\texttt{Taxonomic supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{ taxa}

### 2018-05-07 doc/incertae_sedis.tex
figure 4
figure 5

However, there are cases where thus procedure does not result in one name per
node.  This case can occur, for example, when two \emph{incertae sedis} siblings are 
``inter-digitated'', or  combined into a single clade in such a way that their
members cannot be separated.  When the members of the two clades share the same
MRCA, as in Fig. \ref{fig:interdigitation}, then the names for both clades
apply to the same node.  However, in this case, neither name is ancestral to
the other, and the previous solution does not apply.  In such cases, we choose
one of the names arbitrarily.

@@

\subsection{Placement}\label{subsec:Placement}

@@

A slightly more general case is shown in 
Figure \ref{fig:Broken-incertae-sedis}.
Here taxon $C1$ is an descendant of $A$ in the
synthesis tree, but not the taxonomy.
Thus $C1$ is placed within $A$.
We note that the most recent common ancestor (MRCA) of $A$ and $C1$ in
the taxonomy is the root node, and the path to the MRCA from $A$ is
$A\to AB\to root$, while the path to the MRCA from $C1$ is $C1\to C\to
root$.
Here the additional node $AB$ on the path to the MRCA from $A$
indicates that $C1$ is placed within $AB$ as well as $A$.
In contrast,
the additional node $C$ on the path to the MRCA indicates that $C1$ is
a broken \emph{incertae sedis} taxon.
Thus we may consider an
algorithm that, for each taxon $B$ on the synthesis tree, finds the
closest ancestral taxon $A$, and checks if $A$ is an ancestor of $B$
on the taxonomy.
If not, we compute the paths $A\to A_{1}\to\ldots\to
MRCA(A,B)$ and $B\to B_{1}\to\ldots\to MRCA(A,B)$.

@@

When a clade is marked as \emph{incertae sedis}, we need to consider
two cases.
In the first case, the clade may be placed within a sister
clade intact (Figure \ref{fig:Handling-incertae-sedis}).
However, if
the clade is not monophyletic in the synthesis tree, then we allow the
members of the clade to be placed separately (Figure \ref{fig:Broken-incertae-sedis}).
\begin{figure}
\hfill{}\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-clade/tree1-shaded}

\caption{\label{fig:Broken-incertae-sedis} \emph{Incertae sedis} clade $C$
broken by conflicting placement.
Input trees (a) and (b) conflict
in the placement of taxa in $C$.
The synthesis supertree (d) places
places $C1$ and $C2$ separately within $A$, while unplaced taxa
float upwards.
(\textbf{Note:} This is what the sub-problem solver
does.
The entire pipeline would attach $C3$ at $A$ actually, since
it is the MRCA of $A1$ and $A2$.
Should we change this?) }

@@

In Figure \ref{fig:Broken-incertae-sedis}, tree $T_{1}$ places $C1$
next to $A1$, while $T_{2}$ places $C2$ next to $A2$.
Since different
members of $C$ are placed, neither placement for $C$ is rejected.
Instead the taxon $C$ is broken, the name $C$ disappears, and $C3$
floats to the top level.
Furthermore, since taxon $C$ is a broken
\emph{incertae sedis }taxon, all of its children are effectively
incertae sedis independently, with the exception that they cannot be
placed within each other.
Therefore, the names $A$ and $AB$ are not
lost, since the taxa placed within them are \emph{incertae sedis}
names.

The situation here would be different if the monophyly of $C$ was
supported by an high-ranked input phylogeny.
In that case, instead of
breaking $C$, we would choose the placement in the highest-ranked
tree.

>>>>

However, there are cases where thus procedure does not result in one name per
node.  This case can occur, for example, when two \emph{incertae sedis} siblings are 
``inter-digitated'', or  combined into a single clade in such a way that their
members cannot be separated.  When the members of the two clades share the same
MRCA, as in Fig. \ref{fig:interdigitation}, then the names for both clades
apply to the same node.  However, in this case, neither name is ancestral to
the other, and the previous solution does not apply.
In such cases, we treat all applicable names as synonyms for the node.
\mthcomment{Need to verify that the namer records this situation as synonyms!}

@@

\subsection{Taxonomic revision}\label{subsec:Placement}

@@

A slightly more general case is shown in 
Figure \ref{fig:Broken-incertae-sedis}.
Here taxon $T1$ is an descendant of $R$ in the
synthesis tree, but not the taxonomy.
Thus $T1$ is placed within $R$.
We note that the MRCA of $R$ and $T1$ in
the taxonomy is the root node, and the path to the MRCA from $R$ is
$R\to RS\to root$, while the path to the MRCA from $T1$ is $T1\to T\to
root$.
Here the additional node $RS$ on the path to the MRCA from $R$
indicates that $T1$ is placed within $RS$ as well as $R$.
In contrast,
the additional node $T$ on the path to the MRCA indicates that $T1$ is
a broken \emph{incertae sedis} taxon.
Thus we may consider an
algorithm that, for each taxon $S$ on the synthesis tree, finds the
closest ancestral taxon $A$, and checks if $A$ is an ancestor of $B$
on the taxonomy.
If not, we compute the paths $A\to A_{1}\to\ldots\to
MRCA(A,B)$ and $B\to B_{1}\to\ldots\to MRCA(A,B)$.

\begin{figure}
\hfill{}\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-clade/tree1-shaded}

\caption{\label{fig:Broken-incertae-sedis} \emph{Incertae sedis} clade $T$
broken by conflicting placement.
Input trees (a) and (b) conflict
in the placement of taxa in $T$.
The synthesis supertree (d) places
places $T1$ and $T2$ separately within $R$, while unplaced taxa
float upwards.
}

@@

When a clade is marked as \emph{incertae sedis}, we need to consider
two cases.
In the first case, the clade may be placed within a sister
clade intact (Figure \ref{fig:Handling-incertae-sedis}).
However, if
the clade is not monophyletic in the synthesis tree, then we allow the
members of the clade to be placed separately (Figure \ref{fig:Broken-incertae-sedis}).

@@

In Figure \ref{fig:Broken-incertae-sedis}, tree $T_{1}$ places $T1$
next to $R1$, while $T_{2}$ places $T2$ next to $R2$.
Since different
members of $C$ are placed, neither placement for $T$ is rejected.
Instead the taxon $T$ is broken, the name $T$ disappears, and $T3$
floats to the MRCA of the placed members of $T$.
Furthermore, since taxon $T$ is a broken
\emph{incertae sedis} taxon, all of its children are effectively
incertae sedis independently, with the exception that they cannot be
placed within each other.
Therefore, the names $R$ and $RS$ are not
lost, since the taxa placed within them are \emph{incertae sedis}
names.

The situation here would be different if the monophyly of $T$ was
supported by an high-ranked input phylogeny.
In that case, instead of
breaking $T$, we would choose the placement in the highest-ranked
tree.
